---
title: "Survival Analysis in R"
author: "Andi"
Last updated: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE}
knitr::read_chunk('survival.R')
```

```{r}
# {r, echo = FALSE, results='hide'}
# if we used both 'echo=TRUE' and 'results=hide' the pipe would not work properly
# if we used 'echo = FALSE' and 'results=hide' we would have only messages (i.e. attaching package) If we don't want them we set 'error = FALSE', 'warning = FALSE', and 'message = FALSE'.
library(dplyr)
```

```{r}
# package for analysis
library(survival)

# package for visualization
library(survminer)

# time to death of 686 breast cancer patients
data(GBSG2, package = "TH.data")

# time to re-employment of 3343 unemployed patients
data(UnempDur, package = "Ecdat")
```

# Introducing the GBSG2 dataset

```{r}
# Check out the help page for this dataset
# help(GBSG2, package = "TH.data")

# Look at the summary of the dataset
summary(GBSG2)
```

## Digging into the GBSG2 dataset 1


```{r}
# Count censored and uncensored data
num_cens <- table(GBSG2$cens)
num_cens

# Create barplot of censored and uncensored data
barplot(num_cens)
```

## Using the Surv() function for GBSG2

```{r}
# Create Surv-Object
sobj <- Surv(GBSG2$time, GBSG2$cens)

# Look at 10 first elements
sobj[1:10]
GBSG2$time[1:10]
GBSG2$cens[1:10]

# Look at summary
summary(sobj)

# Look at structure
str(sobj)
```

## The UnempDur dataset

```{r}
# Load the UnempDur data
data(UnempDur, package = "Ecdat")

# Count censored and uncensored data
cens_employ_ft <- table(UnempDur$censor1)
cens_employ_ft

# Create barplot of censored and uncensored data
barplot(cens_employ_ft)

# Create Surv-Object
sobj <- Surv(UnempDur$spell, UnempDur$censor1)

# Look at 10 first elements
sobj[1:10]
```

## First Kaplan-Meier estimate

```{r}
# Create time and event data
time <- c(5, 6, 2, 4, 4)
event <- c(1, 0, 0, 1, 1)

# Compute Kaplan-Meier estimate
km <- survfit(Surv(time, event) ~ 1)
km

# Take a look at the structure
str(km)

# Create data.frame
data.frame(time = km$time, n.risk = km$n.risk, n.event = km$n.event,
  n.censor = km$n.censor, surv = km$surv)
```

## Exercise ignoring censoring

```{r}

# Create dancedat data
dancedat <- data.frame(
  name = c("Chris", "Martin", "Conny", "Desi", "Reni", "Phil", 
    "Flo", "Andrea", "Isaac", "Dayra", "Caspar"),
  time = c(20, 2, 14, 22, 3, 7, 4, 15, 25, 17, 12),
  obs_end = c(1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 0))

# Estimate the survivor function pretending that all censored observations are actual observations. The Surv() function can also take only one argument if there is no censoring, i.e. Surv(time)
km_wrong <- survfit(Surv(time)~ 1, data = dancedat)

# Estimate the survivor function from this dataset via kaplan-meier.
km <- survfit(Surv(time, obs_end) ~ 1, data = dancedat)

# Plot the two and compare
ggsurvplot_combine(list(correct = km, wrong = km_wrong))
```

## Estimating and visualizing a survival curve

```{r}
# Kaplan-Meier estimate
km <- survfit(Surv(time, cens) ~ 1, data = GBSG2)

# plot of the Kaplan-Meier estimate
ggsurvplot(km)

# add the risk table to plot
ggsurvplot(km, risk.table = TRUE)

# add a line showing the median survival time
ggsurvplot(km, risk.table = TRUE, surv.median.line = "hv")
```

## Estimating median survival from a Weibull model

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Compute the median survival from the model
predict(wb, type = "quantile", p = 0.5, newdata = data.frame(1))
```

## Survival curve quantiles from a Weibull model

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# 70 Percent of patients survive beyond time point...
predict(wb, type = "quantile", p = 1- 0.7, newdata = data.frame(1))
```

## Estimating the survival curve with survreg()

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Retrieve survival curve from model probabilities 
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1 - surv, newdata = data.frame(1))

# Create data frame with the information
surv_wb <- data.frame(time = t, surv = surv)

# Look at first few lines of the result
head(surv_wb)
```

## Comparing Weibull model and Kaplan-Meier estimate

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Retrieve survival curve from model
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1 - surv, newdata = data.frame(1))

# Create data frame with the information needed for ggsurvplot_df
surv_wb <- data.frame(time = t, surv = surv, 
  upper = NA, lower = NA, std.err = NA)

# Plot
ggsurvplot_df(fit = surv_wb, surv.geom = geom_line)
```

## Interpreting coefficents

```{r, interp coef, echo=FALSE}
# load a new data set of lung cancer patients
```

```{r}
# Look at a new data set of lung cancer patients
str(dat)

# Estimate a Weibull model
wbmod <- survreg(Surv(time, status) ~ sex, data = dat)
coef(wbmod)
```

## Compute Weibull model

```{r}
# Weibull model
wbmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2)

# Retrieve survival curve from model
surv <- seq(.99, .01, by = -.01)
t_yes <- predict(wbmod, type = "quantile", p = 1 - surv,
  newdata = data.frame(horTh = "yes"))

# Take a look at survival curve
str(t_yes)
```

Note that if we put `newdata = data.farme(horTh = 2)` in the above code, we would get an error.

## Computing a Weibull model and the survival curves

```{r}
# Weibull model
wbmod <- survreg(Surv(time, cens) ~ horTh + tsize, data = GBSG2)

# Imaginary patients
newdat <- expand.grid(
  horTh = levels(GBSG2$horTh),
  tsize = quantile(GBSG2$tsize, probs = c(0.25, 0.5, 0.75)))

# Compute survival curves
surv <- seq(.99, .01, by = -.01)
t <- predict(wbmod, type = "quantile", p = 1 - surv,
  newdata = newdat)

# How many rows and columns does t have?
dim(t)

# Each row of t corresponds to one covariate combination (one imaginary patient) and each column to one value of surv.
```

## Visualising a Weibull model

```{r}
# Use cbind() to combine the information in newdat with t
surv_wbmod_wide <- cbind(newdat, t)
  
# Use melt() to bring the data.frame to long format
library(reshape2)
surv_wbmod <- melt(surv_wbmod_wide, id.vars = c("horTh", "tsize"), variable.name = "surv_id", value.name = "time")

# Use surv_wbmod$surv_id to add the correct survival probabilities surv
surv_wbmod$surv <- surv[as.numeric(surv_wbmod$surv_id)]

# Add columns upper, lower, std.err, and strata to the data.frame
surv_wbmod[, c("upper", "lower", "std.err", "strata")] <- NA

# Plot the survival curves
ggsurvplot_df(surv_wbmod, surv.geom = geom_line,
  linetype = "horTh", color = "tsize", legend.title = NULL)
```

## Computing a Weibull and a log-normal model

```{r}
# Weibull model
wbmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2)

# Log-Normal model
lnmod <- survreg(Surv(time, cens) ~ horTh, data = GBSG2, dist = "lognormal")

# Newdata
newdat <- data.frame(horTh = levels(GBSG2$horTh))

# Surv
surv <- seq(.99, .01, by = -.01)

# Survival curve from Weibull model and log-normal model
wbt <- predict(wbmod, type = "quantile", p = 1 - surv, newdata = newdat)
lnt <- predict(lnmod, type = "quantile", p = 1 - surv, newdata = newdat)
```

## Comparing Weibull and Log-Normal Model

```{r}
# Create a wide data frame containing hormonal therapy information and the survival curves for the Weibull and log-normal models.
surv_wide <- rbind(cbind(newdat, wbt), cbind(newdat, lnt))
surv_wide$dist <- c("weibull", "weibull", "lognormal", "lognormal")

surv_wide[, 1:5]
surv_wide[, ncol(surv_wide) - 4]

# Melt the data.frame into long format.
surv_long <- melt(surv_wide, id.vars = c("horTh", "dist"), variable.name = "surv_id", value.name = "time")

# Add column for the survival probabilities
surv_long$surv <- surv[as.numeric(surv_long$surv_id)]

# Add columns upper, lower, std.err, and strata contianing NA values
surv_long[, c("upper", "lower", "std.err", "strata")] <- NA

# Plot the survival curves
ggsurvplot_df(surv_long, surv.geom = geom_line,
              linetype = "horTh", color = "dist", legend.title = NULL)
```

## Computing a Cox model

```{r, eval=FALSE}
# Compute Cox model
cxmod <- coxph(Surv(time, status) ~ performance, data = dat)

# Show model coefficient
coef(cxmod)
```

## Computing the survival curve from a Cox model

```{r}
# Cox model
cxmod <- coxph(Surv(time, cens) ~ horTh + tsize, data = GBSG2)

# Imaginary patients
newdat <- expand.grid(
  horTh = levels(GBSG2$horTh),
  tsize = quantile(GBSG2$tsize, probs = c(0.25, 0.5, 0.75)))
rownames(newdat) <- letters[1:6] # Those are stratum names, we need them.

# explore newdat
newdat

# Compute survival curves
cxsf <- survfit(cxmod, data = GBSG2, newdata = newdat, conf.type = "none")

# look at he structure of cxsf
str(cxsf)

# What does cxsf look ike? (not very useful!)
cxsf

# Look at first 6 rows of cxsf$surv and time points
head(cxsf$surv)
head(cxsf$time)
```

## Visualizing a Cox model

`surv_summary()` is a function that returns a data.frame containing a nice summary from `survfit()` results, including columns like time (survival time) and surv (survival probability).

```{r}
# Compute data.frame needed for plotting
surv_cxmod0 <- surv_summary(cxsf)
str(surv_cxmod0)
dim(surv_cxmod0)
head(surv_cxmod0)
tail(surv_cxmod0)

# Get a character vector of patient letters (patient IDs)
pid <- as.character(surv_cxmod0$strata)
length(pid)
head(pid, 12)
tail(pid, 12)

# Multiple of the rows in newdat so that it fits with surv_cxmod0
m_newdat <- newdat[pid, ]
dim(m_newdat)
head(m_newdat)
tail(m_newdat)

# Add patient info to data.frame
surv_cxmod <- cbind(surv_cxmod0, m_newdat)

# Plot
ggsurvplot_df(surv_cxmod, linetype = "horTh", color = "tsize",
  legend.title = NULL, censor = FALSE)
```

## Capstone: The Cox model

```{r, capstone, echo=FALSE}

```

```{r}
# Compute Cox model and survival curves
cxmod <- coxph(Surv(time, status) ~ performance, data = lung)
new_lung <- data.frame(performance = c(60, 70, 80, 90))
cxsf <- survfit(cxmod, data = lung, newdata = new_lung, conf.type = "none")

# Use the summary of cxsf to take a vector of patient IDs
surv_cxmod0 <- surv_summary(cxsf)
pid <- as.character(surv_cxmod0$strata)

# Duplicate rows in newdat to fit with surv_cxmod0 and add them in
m_newdat <- new_lung[pid, , drop = FALSE]
surv_cxmod <- cbind(surv_cxmod0, m_newdat)

# Plot
ggsurvplot_df(surv_cxmod, color = "performance", legend.title = NULL, censor = FALSE)
```

## Capstone: Comparing survival curves

```{r}
# Compute Kaplan-Meier curve
km <- survfit(Surv(time, status) ~ 1, data = lung)

# Compute Cox model
cxmod <- coxph(Surv(time, status) ~ performance, data = lung)

# Compute Cox model survival curves
new_lung <- data.frame(performance = c(60, 70, 80, 90))
cxsf <- survfit(cxmod, data = lung, newdata = new_lung, conf.type = "none")

# Plot Kaplan-Meier curve
ggsurvplot(km, conf.int = FALSE)

# Plot Cox model survival curves
ggsurvplot(cxsf, censor = FALSE)
```

## Where do we go from here?

Check out CRAN task view.