---
title: "Survival Analysis in R"
author: "Andi"
Last updated: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE}
knitr::read_chunk('survival.R')
```

```{r}
# {r, echo = FALSE, results='hide'}
# if we used both 'echo=TRUE' and 'results=hide' the pipe would not work properly
# if we used 'echo = FALSE' and 'results=hide' we would have only messages (i.e. attaching package) If we don't want them we set 'error = FALSE', 'warning = FALSE', and 'message = FALSE'.
library(dplyr)
```

```{r}
# package for analysis
library(survival)

# package for visualization
library(survminer)

# time to death of 686 breast cancer patients
data(GBSG2, package = "TH.data")

# time to re-employment of 3343 unemployed patients
data(UnempDur, package = "Ecdat")
```

# Introducing the GBSG2 dataset

```{r}
# Check out the help page for this dataset
# help(GBSG2, package = "TH.data")

# Look at the summary of the dataset
summary(GBSG2)
```

## Digging into the GBSG2 dataset 1


```{r}
# Count censored and uncensored data
num_cens <- table(GBSG2$cens)
num_cens

# Create barplot of censored and uncensored data
barplot(num_cens)
```

## Using the Surv() function for GBSG2

```{r}
# Create Surv-Object
sobj <- Surv(GBSG2$time, GBSG2$cens)

# Look at 10 first elements
sobj[1:10]
GBSG2$time[1:10]
GBSG2$cens[1:10]

# Look at summary
summary(sobj)

# Look at structure
str(sobj)
```

## The UnempDur dataset

```{r}
# Load the UnempDur data
data(UnempDur, package = "Ecdat")

# Count censored and uncensored data
cens_employ_ft <- table(UnempDur$censor1)
cens_employ_ft

# Create barplot of censored and uncensored data
barplot(cens_employ_ft)

# Create Surv-Object
sobj <- Surv(UnempDur$spell, UnempDur$censor1)

# Look at 10 first elements
sobj[1:10]
```

## First Kaplan-Meier estimate

```{r}
# Create time and event data
time <- c(5, 6, 2, 4, 4)
event <- c(1, 0, 0, 1, 1)

# Compute Kaplan-Meier estimate
km <- survfit(Surv(time, event) ~ 1)
km

# Take a look at the structure
str(km)

# Create data.frame
data.frame(time = km$time, n.risk = km$n.risk, n.event = km$n.event,
  n.censor = km$n.censor, surv = km$surv)
```

## Exercise ignoring censoring

```{r}

# Create dancedat data
dancedat <- data.frame(
  name = c("Chris", "Martin", "Conny", "Desi", "Reni", "Phil", 
    "Flo", "Andrea", "Isaac", "Dayra", "Caspar"),
  time = c(20, 2, 14, 22, 3, 7, 4, 15, 25, 17, 12),
  obs_end = c(1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 0))

# Estimate the survivor function pretending that all censored observations are actual observations. The Surv() function can also take only one argument if there is no censoring, i.e. Surv(time)
km_wrong <- survfit(Surv(time)~ 1, data = dancedat)

# Estimate the survivor function from this dataset via kaplan-meier.
km <- survfit(Surv(time, obs_end) ~ 1, data = dancedat)

# Plot the two and compare
ggsurvplot_combine(list(correct = km, wrong = km_wrong))
```

## Estimating and visualizing a survival curve

```{r}
# Kaplan-Meier estimate
km <- survfit(Surv(time, cens) ~ 1, data = GBSG2)

# plot of the Kaplan-Meier estimate
ggsurvplot(km)

# add the risk table to plot
ggsurvplot(km, risk.table = TRUE)

# add a line showing the median survival time
ggsurvplot(km, risk.table = TRUE, surv.median.line = "hv")
```

## Estimating median survival from a Weibull model

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Compute the median survival from the model
predict(wb, type = "quantile", p = 0.5, newdata = data.frame(1))
```

## Survival curve quantiles from a Weibull model

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# 70 Percent of patients survive beyond time point...
predict(wb, type = "quantile", p = 1- 0.7, newdata = data.frame(1))
```

## Estimating the survival curve with survreg()

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Retrieve survival curve from model probabilities 
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1 - surv, newdata = data.frame(1))

# Create data frame with the information
surv_wb <- data.frame(time = t, surv = surv)

# Look at first few lines of the result
head(surv_wb)
```

## Comparing Weibull model and Kaplan-Meier estimate

```{r}
# Weibull model
wb <- survreg(Surv(time, cens) ~ 1, data = GBSG2)

# Retrieve survival curve from model
surv <- seq(.99, .01, by = -.01)

# Get time for each probability
t <- predict(wb, type = "quantile", p = 1 - surv, newdata = data.frame(1))

# Create data frame with the information needed for ggsurvplot_df
surv_wb <- data.frame(time = t, surv = surv, 
  upper = NA, lower = NA, std.err = NA)

# Plot
ggsurvplot_df(fit = surv_wb, surv.geom = geom_line)
```

## Interpreting coefficents

```{r, interp coef, echo=FALSE}
# load a dataset of lung cancer patients
```

